Digital economy in the UK: a multi-scalar story of the diffusion of web technologies

### 2023-10-27 paper/submission1/tranos2023.qmd
presentation

Geographers were always interested in how new technologies and
innovations diffuse across space and, importantly, how such
spatio-temporal processes can be modelled. After all, diffusion together 
with invention and innovation are considered the pillars of 
technological change [@das2022diffusion]. The seminal contribution of
@hagerstrand1968innovation is illustrative of this early interest.
However, the torch of exploring and modelling such processes had been
passed to other disciplines such as economics, business studies and
sociology well before the 'cultural turn' of economic geography
[@perkins2005international]. A potential explanation of the lack of 
geographical studies exploring the diffusion of new and, more 
specifically, digital technologies across *both* space and time can be
attributed to the scarcity of relevant and granular enough data. As 
@zook2022mapping highlight, digital activities are hardly ever captured 
in official data.

This paper offers such a contribution: a geographical study illustrating 
how a new technology that is the web diffused over space and time 
in the UK at a high level of spatial granularity during the $1996-2012$ 
period. It does so by employing a novel source of big data which 
captures the active engagement with web technologies during that period.
By addressing this empirical question this paper exemplifies how the 
combination of data sources which escape the traditional social science
domain and adequate research methods can offer new contributions 
regarding our understanding of technologies are diffused.

The motivation for this paper lies in the fact that there are various
stakeholders who are interested in knowing how new digital technologies
diffused over space and time and use this information to make predictions
regarding the diffusion of related *future* technologies.
As per @leibowicz2016representing, historical studies agree on the fact that 
"technologies diffuse at different times, at different rates, and to
different extents in different places, and can be significantly
influenced by policies [@victor1993]". @meade2021modelling highlight that 
a variety of actors have a direct interest in gaining such knowledge 
including network equipment suppliers; network operators, regulatory 
and local authorities. 
These processes and their effects very a lot between scales: although
the diffusion of a new technology might not be optimal at a local level,
it might be beneficial from a global perspective as it might enable 
faster diffusion to less advantaged places @leibowicz2016representing. 
Despite the spatial heterogeneity of such diffusion mechanisms and 
the policy relevance, there are very limited attempts in the literature 
to analyse the diffusion of new digital technologies at a detailed 
geographical level.

<!-- diffusion and adoption -->

Technological diffusion, which is by definition an aggregated process, 
can be discussed in parallel with individual adoption mechanisms.
On the one hand, @rogers2010diffusion identifies early adopter of new 
technologies as 'knowledgeable risk takers' and @griliches1957 as 'profit 
maximisers' [@ding2010modeling]. Such individual agents are rewarded because 
of their attitude towards new technologies and innovations. On the other hand,
@perkins2011internet attribute diffusion to two processes: (i) epidemic-type 
of mechanisms, which

Two main mechanisms are identified in the
literature to explain diffusion: (1) epidemic-type dynamics whereby
contact with previous adopters stimulates uptake as po tential adopters
learn about a new innovation; and (2) economic-type mechanisms whereby
potential users adopt a new innovation as it be comesmore profitable,
useful, or valuable, with uptake characteristically spreading as costs
be come lower, performance improves, or the po tential uses of the
innovation grow over time.
The criticisms are mainly based on the fact that although the approach
gives an idea of aggregate (industry or household) behaviour, it does
not focus on the individual's (firm or household) adoption process

@morrill2020spatial understands spatial diffusion as the landscape 
changes that spread across space because of changes that take place 
elsewhere.

<!-- data and early engagement with the web -->

wilson201281 Early on in their lifecycle, new technologies are crude,
imperfect, and expensive (Rosenberg, 1994). New energy technologies are
attractive for their ability to perform a particular task or deliver a
new or improved energy service (Fouquet, 2010). This is often
circumscribed by a particular set of needs in a particular context: a
market 'niche'. End-users in niche markets are generally less sensitive
to the effective price of the energy service provided or have a higher
willingness to pay for its performance advantages (Fouquet, 2010). Thus
initially, performance dominates cost competitiveness (Wilson and
Grubler, 2011). Market niches afford some protection from competitive
pressures, allowing technologies to be tested and improved in applied
settings, reducing uncertainties with performance or market demand (Kemp
et al., 1998). Costs may only fall substantively after an extended
period of commercial experimentation, concurrent with the establishment
of an industrial base and characteristic moves towards standardisation
and mass production (Grubler, 1998). The influence of accumulating
production experience on costs is captured by the concept of learning.

@perkins2005international Theoretical models disagree as to why firms
adopt innovations at different times. Epidemic models emphasize
information (Griliches 1957). Certain firms are hypothesized to adopt
earlier because they come into contact with, and learn from, adopters of
the new technology before others. Economic models, on the other hand,
predominantly emphasize firm heterogeneity (Ireland and Stoneman 1986).
Firms adopt technologies at different times because they differ with
respect to various organizational and environmental variables
influencing the economic returns from adoption (Blackman 1999).

<!-- methods -->

@grubler1990rise Later Hagerstrand conceptualized physical "barrier"
effects like lakes or uninhabited areas, which, in addition to distance,
act as further retarding effects on diffusion. These are formalized in
the form of "zero" or "half" contact multiplicators on the (distance
decaying) message flows.

@@

<!-- methods -->

@@

MAYBE SKIP THIS??

<!-- previous studies -->

@beardsell1999spatial To characterize evolution of the computer
industry, we examine the distribution of relative employment across
cities in 1977 and how that distribution changes over time to 1992

There is no tendency of relative size distributions of urban computer
employment to collapse, go bimodal, or fully spread. Overall computers
exhibit some turbulence, with dramatic big winners and losers among
cities. In attracting or repelling an industry, urban heterogeneity is
important. Large, well educated cities near San Jose have much greater
chances of attracting high-tech employment and less of losing it

@bednarz2020pulled spatial innovation diffusion By using Bayesian
survival models with time-dependent data of wind turbine deployment and
firm foundation for 402 German regions between the years 1970 and 2015,
we show that the spatial evolution of the German wind energy industry
was more strongly influenced by local demand-- pull than local
supply--push processes.

The industry's initial locations are distributed relatively arbitrarily
and unpredictably, as their needs in terms of resources and skills are
diverse and distinct from the older existing industries (Boschma and
Lambooy, 1999). Consequently, emerging industries are characterized by
relatively high degrees of freedom in terms of location. In later
extensions of the concept, the assumption of the randomness of locations
was revised with greater importance assigned to regional conditions
(Boschma and Lambooy, 1999; Fornahl et al., 2012).

@perkins2005international analyze whether the rate at which new producer
technologies diffuse is significantly influenced by (1) latecomer
advantage and (2) engagement with the global economy via trade and
foreign investment

Indeed, precisely because of these latecomer advantages, developing
countries1 are believed to be well placed to catch up with developed
ones (Gerschenkron 1962; Abramovitz 1986).

@fritsch2015new analyze the spatial diffusion of laser technology
research in West Germany from 1960, when this technology began, until
2005.

@perkins2011internet

@lengyel2020role We see that the OSN spread almost exclusively from the
original location (the capital Budapest, with an order of magnitude more
inhabitants than the next size town) to various parts of the country in
the early phase of the life-cycle. Later, diffusion became less
mono-centric and other towns also emerged as spreaders. Our findings
support the idea that spreading initially happens to large distances and
becomes more local over time. This is illustrated and discussed later in
Figure 2I.

@PAPAGIANNIDIS2015308

@leibowicz2016representing for energy There is no generally accepted
theory that explains diffusion rate heterogeneity across technologies,
but several factors are considered important. Greater unit scale and
larger market size contribute to slower diffusion. Requirements for
interrelated technologies or complex infrastructures also hinder the
diffusion process (Grubler, 2012).

Mobile phones benefited from early deployment in recreational boats and
automobiles, where the traditional competitor was not a viable option.
In the early stages of diffusion, performance is a more important driver
of adoption than cost competitiveness. Typically, significant cost
reductions only occur once the technology reaches a deployment level
capable of supporting standardization and mass production (Wilson,
2012).

@leibowicz2016representing Empirical evidence supports the validity of
Schmidt's Law over a wide range of technologies, time periods, and
geographical contexts. A recent meta-analysis of technology up-scaling
found that diffusion accelerated moving from the core to the rim and
periphery for technologies as diverse as natural gas power, oil
refineries, and automobiles (Wilson, 2009). One historical example that
conforms particularly well to Schmidt's Law is the diffusion of coal
power in Europe (Grubler, 2012). England emerged as the core region for
coal power because it had legal and economic institutions that
incentivized scientific pursuits, domestic coal reserves, and a clear
industrial motivation to replace water power with coal. ...

@ding2010modeling model the spatial diffusion of mobile
telecommunications in China as well as its determinants. regional
socioeconomic characteristics play an important role in determining the
timing, speed, and level of mobile telecommunications diffusion in China

@bento2018time explore What determines the duration of formative phases
for energy innovations in different markets? We are interested both in
initial markets (also: core, lead, first mover, early adopter) where
formative phases prepare technologies for mass commercialization, and in
follower markets (also: periphery, lag, late adopter) where accelerated
formative phases may benefit from diffusion and spillovers.

@perkins2005internationalEither way, the strong assumption is that developing countries can
acquire modern technology innovated in developed economies, often at a
fraction of the original research and development (R&D) costs, thereby
leapfrogging many decades of technological progress (Teece 2000).

@perkins2005international The first involves a country's geographical location. Recent empirical
work suggests that diffusion is ''geographically localized'' (Globerman,
Kokko, and Sjo¨holm 2000; Keller 2002; Milner 2003) in that a technology
diffuses faster in a country where it is already more widely diffused in
neighboring countries. Underlying these regional effects are contagion
and contact with prior users or producers of technology.

@@

>>>>

<!-- geographers lack of interest + motivation -->
Geographers were always interested in how new technologies and
innovations diffuse across space and, importantly, how such
spatio-temporal processes can be modelled. After all, diffusion together 
with invention and innovation are considered the pillars of 
technological change [@das2022diffusion]. The seminal contribution of
@hagerstrand1968innovation is illustrative of this early interest.
However, the torch of exploring and modelling such processes had been
passed to other disciplines such as economics, business studies and
sociology well before the 'cultural turn' of economic geography
[@perkins2005international]. A potential explanation of the lack of 
geographical studies exploring the diffusion of new and, more 
specifically, digital technologies across *both* space and time can be
attributed to the scarcity of relevant and granular enough data. As 
@zook2022mapping highlight, digital activities are hardly ever captured 
in official data.

<!-- contribution -->
This paper offers such a contribution: a geographical study illustrating 
how a new technology that is the web diffused over space and time 
in the UK at a high level of spatial granularity during the $1996-2012$ 
period. It does so by employing a novel source of big data which 
captures the active engagement with web technologies during that period.
By addressing this empirical question this paper exemplifies how the 
combination of data sources which escape the traditional social science
domain and adequate research methods can offer new contributions 
regarding our understanding of how new technologies diffuse.

<!-- motivation -->
The motivation for this paper lies in the fact that there are various
stakeholders who are interested in knowing how new digital technologies
diffused over space and time and use this knowledge to make predictions
regarding the diffusion of related *future* technologies.
As per @leibowicz2016representing, historical studies agree on the fact that 
"technologies diffuse at different times, at different rates, and to
different extents in different places, and can be significantly
influenced by policies [@victor1993]". @meade2021modelling highlight that 
a variety of actors have a direct interest in gaining such knowledge 
including network equipment suppliers; network operators, regulatory 
and local authorities. 
These processes and their effects very a lot between scales: although
the diffusion of a new technology might not be optimal at a local level,
it might be beneficial from a global perspective as it could lead to 
faster diffusion to less advantaged places [@leibowicz2016representing]. 
Despite the spatial heterogeneity of such diffusion mechanisms and 
the policy relevance, there are very limited attempts in the literature 
to analyse the diffusion of new digital technologies at a detailed 
geographical level.

<!-- diffusion and adoption -->
Technological diffusion, which is by definition an aggregated process, 
can be discussed in parallel with individual adoption mechanisms.
On the one hand, @rogers2010diffusion identifies early adopter of new 
technologies as 'knowledgeable risk takers' and @griliches1957 as 'profit 
maximisers' [@ding2010modeling]. Such individual agents are rewarded because 
of their attitude towards new technologies and innovations. On the other hand,
@perkins2011internet attribute diffusion to two processes: (i) epidemic-like 
mechanisms, which are governed by distance, proximity and social interactions, 
and (ii) by economic mechanisms as new innovations are adopted by users as they
become more profitable, valueable and useful.
<!-- The criticisms are mainly based on the fact that although the approach -->
<!-- gives an idea of aggregate (industry or household) behaviour, it does -->
<!-- not focus on the individual's (firm or household) adoption process -->

<!-- data and early engagement with the web -->
This paper focuses on the diffusion of the web as new technology during the $1996-2012$
period. This was an exciting period for digital technologies as it corresponds with
the commercialisation of the internet and, consequently, its almost universal 
adoption. The reader is reminded that it was only in 1994 when Netscape Navigator 
was introduced, a year before Microsoft's Internet Explorer
^[[https://www.theguardian.com/global/2015/mar/22/web-browser-came-back-haunt-microsoft](https://www.theguardian.com/global/2015/mar/22/web-browser-came-back-haunt-microsoft)].
Also, only $9$ per cent of UK's households had access to the internet in $1998$ 
[@ons2018], the web included mostly static webpages, there were no social media 
and web browsing was happening exclusively from desktop PCs as there were no 
smartphones [@tranosuk]. 
Hence, it is fair to say that the study period captures the very early stages of 
the diffusion on a new technology that is the web. 
This is key point in the lifecycle of a new technology. Firstly, during this period
new technologies are expensive, crude and imperfect [@rosenberg1994exploring; @wilson201281].
A simple comparison between Web 1.0 and Web 3.0 applications clearly illustrates this
[@tranos2020social]. During this period the performance of a new technology
is the main attraction and not the cost to access and use it [@wilson2011lessons]
There is a broader theoretical discussion in the literature regarding the motivation
behind early adoption. As summarised by @perkins2005international, on the one hand,
epidemic models highlight the role of interpersonal contacts as a way for new 
technologies to diffuse. On the other hand, economic models underline the importance
of heterogeneity. Different firms have different structures and business plans,
which define the potential economic returns of the adoption of a new technology 
and, therefore, the choice to adopt a new technology becomes an individual option.
From a broader and evolutionary perspective, initial conditions are essential for 
the creation and evolution of path-dependent technological development trajectories
[@neffke2011regions; @simmie2014new]. This argument is even more relevant when the
focus is on Information and Communication Technologies (ICT) because of the commonly
found lag between investment and economic returns as reflected in the Solow paradox 
[@acemoglu2014return; @brynjolfsson2018artificial].

Importantly, the data used here depicts the *active* engagement with the web as 
it contains all the commercial websites archived by the Internet Archive
^[[https://archive.org/](See https://archive.org/).]
that include a mention to at least one valid UK postcode in the web text. 
Websites are identified as commercial ones if they are part of the .co.uk second 
level domain (SLD). Also, the act of of creating a website is understood here as 
active engagement with the web vis-à-vis the more passive act of browsing the web 
or having an internet connection [@tranosuk]. Previous studies have focused mostly
on more passive notions of engaging with digital technologies such as internet 
adoption and internet speeds [e.g. @blank2018local; @riddlesden2014broadband].

<!-- description of the methods and results -->

<!-- contents -->

<!-- methods -->
@grubler1990rise Later Hagerstrand conceptualized physical "barrier"
effects like lakes or uninhabited areas, which, in addition to distance,
act as further retarding effects on diffusion. These are formalized in
the form of "zero" or "half" contact multiplicators on the (distance
decaying) message flows.

@@

@@

<!-- key concepts -- copied from help.Rmd -->
Geographical diffusion is a synthesis of different processes. On the one hand, we 
can identify purely spatial or, in other words, contagious processes. Adjacency 
and, more broadly, distance are the key drivers of diffusion. 
This perspective draws similarities with epidemics: innovation just like pathogens 
spreads because of contagion and, consequently, proximity and exposure 
[@hivner2003facilitating]. On the other hand, we can identify hierarchical 
processes. Instead of horizontal distance-based diffusion mechanisms, the top-down 
hierarchy of urban systems drives diffusion. In reality, the synthesis of these 
two processes represents how new technologies diffuse over space and time [@morrill2020spatial].

These ideas were firstly introduced by Torsten Hägerstrand and his thesis entitled 
'Innovation Diffusion as a Spatial Process' [@hagerstrand1968innovation]. 
Hägerstrand was the first one to identify diffusion as a geographical process. 
The starting point was the idea that diffusion is based on passing information 
through social networks, which themselves tend to be defined by geography. Hence, 
he identified the 'neighbourhood' effect of how information, and consequently, 
innovation diffuse. He used agricultural innovations to test and model his ideas 
using Monte Carlo simulations. Hägerstrand also incorporated the role of hierarchy 
and how some phenomena maybe firstly adopted in larger cities and then diffuse 
to second tier ones. This is a sequential instead of a simultaneous process, which 
resembles the 'lead-lag' spatial acceleration effect in market research 
[@bento2018time; @PERES201091]. Hägerstrand maybe more widely known for highlighting 
the role time plays in the diffusion of innovations: an early-pioneering 
period, a middle fast accelerating period and a final saturation one [@morrill2020spatial.

The temporal dimension was further explored by Everett Rogers and his seminal work 
on 'Diffusion of Innovations' [@rogers2010diffusion]. Rogers being a sociologist 
his work focused not on the diffusion of innovations over space and time, but 
instead on the adoption of new technologies and innovations by individuals and the 
individual mechanisms that drive the decisions behind adoption. He identified five 
groups of individuals regarding their adoption speed: innovators, early adopters, 
early majority, late majority and laggards. Key mechanism of diffusion and adoption 
is always communication and how knowledge is transferred within a social system. 
Therefore, all approaches agree on the S-shaped diffusion and cumulative adoption 
pattern [@grubler1990rise].

Schmidt's Law empirically illustrates a similar pattern. *Core* and usually highly 
agglomerated regions is where new technologies are invented and commercially deployed 
[@grubler1990rise; grubler1990rise]. This is where the first adopters tend to be 
based. Then, technologies spread to the *rim* and eventually to the *periphery*. 
Although adoption pace might be higher when new technologies finally arrive to the 
periphery, the saturation levels there may never reach the ones in the core because 
of the lack of infrastructure or other necessary institutions [@leibowicz2016representing].
@grubler1990rise effectively summarises the three key characteristics of the spatial 
diffusion process: (i) the cumulative level of adoption follows an S-shaped pattern 
just like purely temporal models; (ii) diffusion is shaped by a hierarchy effect 
in a form of a centrifugal force: from core to periphery; and (iii) diffusion is 
also shaped by distance and a neighbourhood effect and contaminate nearby locations.

<!-- previous studies -->
The remaining of this section reviews empirical studies which analysed the diffusion 
on new technologies over space and time. Although the spatial dimension is present
in most of the following studies, the level of spatial detail is almost always 
more coarse than the one adopted here. To begin with @beardsell1999spatial studied 
the evolution of the computer industry in $317$ US metro areas during the $1977-1992$ 
period using employment data. Their analysis indicated that the relative size distribution
holds for urban computer employment and also urban heterogeneity is essential in
explaining this distribution. @bednarz2020pulled focused on wind turbines and modelled
their spatial diffusion across $402$ German regions during $1970-2015$. Their key
finding is that local demand than local supply was the main driving factor.

At a global scale @perkins2005international analyzed if the diffusion rate of new
technologies is driven by a latecomer advantage and the engagement with the global
economy via foreign direct investments and trade. Their results illustrate that 
indeed latecomers and developing countries experience diffusion of new technologies 
more rapidly than early adopter and developed countries. At the same scale, 
@perkins2011internet explored whether the adoption of previous communication
technologies that is mail, telegrams and telephones was shaped by similar socioeconomic 
factors as the internet. Their results indicated common patterns regarding the 
drivers behind the adoption of different communication technologies.

Turning to studies that share more technological and scalar similarities with this 
paper, @ding2010modeling modelled the spatial diffusion of mobile telecommunications 
across regions in China. Their analysis indicated that socioeconomic characteristics 
are important determinants of the timing, speed and the level of mobile diffusion
within China. Using data from a Hungarian online social network, @lengyel2020role 
analysed its adoption and the churn at a very granular spatial level. Their results 
are in agreement with early theoretical and empirical contributions reviewed here:
assortativity, urban scaling and distance are the key drivers of spatial diffusion.
At a global scale @PAPAGIANNIDIS2015308 modelled the diffusion of different web
technologies technologies and practices. Interestingly, they did so by using similar,
but less extensive data as the one used here. Their analysis illustrated how the 
diffusion of different web technologies and practices follow an S-shaped pattern
as well as the different diffusion rates of the different technologies and practices.

<!-- @fritsch2015new analyze the spatial diffusion of laser technology -->
<!-- research in West Germany from 1960, when this technology began, until -->
<!-- 2005. -->
<!-- in one region  -->

<!-- @leibowicz2016representing for energy There is no generally accepted -->
<!-- theory that explains diffusion rate heterogeneity across technologies, -->
<!-- but several factors are considered important. Greater unit scale and -->
<!-- larger market size contribute to slower diffusion. Requirements for -->
<!-- interrelated technologies or complex infrastructures also hinder the -->
<!-- diffusion process (Grubler, 2012). -->

<!-- Mobile phones benefited from early deployment in recreational boats and -->
<!-- automobiles, where the traditional competitor was not a viable option. -->
<!-- In the early stages of diffusion, performance is a more important driver -->
<!-- of adoption than cost competitiveness. Typically, significant cost -->
<!-- reductions only occur once the technology reaches a deployment level -->
<!-- capable of supporting standardization and mass production (Wilson, -->
<!-- 2012). -->

<!-- @leibowicz2016representing Empirical evidence supports the validity of -->
<!-- Schmidt's Law over a wide range of technologies, time periods, and -->
<!-- geographical contexts. A recent meta-analysis of technology up-scaling -->
<!-- found that diffusion accelerated moving from the core to the rim and -->
<!-- periphery for technologies as diverse as natural gas power, oil -->
<!-- refineries, and automobiles (Wilson, 2009). One historical example that -->
<!-- conforms particularly well to Schmidt's Law is the diffusion of coal -->
<!-- power in Europe (Grubler, 2012). England emerged as the core region for -->
<!-- coal power because it had legal and economic institutions that -->
<!-- incentivized scientific pursuits, domestic coal reserves, and a clear -->
<!-- industrial motivation to replace water power with coal. ... -->

<!-- country level, not sure if needed. -->

<!-- @bento2018time explore What determines the duration of formative phases -->
<!-- for energy innovations in different markets? We are interested both in -->
<!-- initial markets (also: core, lead, first mover, early adopter) where -->
<!-- formative phases prepare technologies for mass commercialization, and in -->
<!-- follower markets (also: periphery, lag, late adopter) where accelerated -->
<!-- formative phases may benefit from diffusion and spillovers. -->

@@

<!-- from our RS paper -->
For instance, we can
approximate digital economic activities in a region by
measuring the volume of commercial webpages by considering
only .co.uk websites, which are dedicated to commercial
activities (Thelwall, 2000). Such commercial
websites are used to exchange information, support online
transactions and share opinions (Blazquez & Domenech,
2018). Although nothing prevents a UK-based company
from adopting a generic TLD such as.com and, indeed,
such cases escape our data, we do not expect that such
omissions could affect our results given the popularity of
the .uk ccTLD: UK consumers prefer to visit a .uk website
when they are searching for products or services (Hope,
2017); and anecdotal evidence indicates that during the
first half of 2000, three .co.uk domains were registered
every minute (OECD, 2001). Also, as Table 1 demonstrates,
.co.uk was the most popular SLD under the .uk
ccTLD in 2000. Hence, we use the total volume of all
the archived webpages in a region as a proxy for the level
of digitization. We start the analysis by focusing on commercial
webpages (.co.uk).
